---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# penguineExplorer

**ETC5523 – Assignment 4: Turning analysis interactive**
`Author: Nishta Bollavaram`
**penguineExplorer** ships a cleaned penguins dataset and a Shiny app to explore
species differences across years. Everything is packaged (no read.csv inside
the app), documented, and published via pkgdown.

## Installation

```{r}
install.packages("remotes")
remotes::install_github(
  "ETC5523-2025/assignment-4-packages-and-shiny-apps-nishtabollavaram"
)
library(penguinExplorer)

```

## Launching the Shiny app

```{r }
run_app()
```

The interactive Shiny explorer allows you to: Select a specie, choose a numeric variables for X and Y axes, compare male and female penguins visually and download filtered data as a CSV.rmarkdown::render("README.Rmd")
pkgdown::build_site()


```{r pressure, echo = FALSE}
head(penguins)

```

## Acknowledgements
Data adapted from the palmerpenguins project.
