---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  message = FALSE,
  warning = FALSE
)
```

# Explore Penguin Measurements with Shiny

**Author:** Nishta Bollavaram  
**Unit:** ETC5523 — *Communicating with Data (Assignment 4)*  
**University:** Monash University  

---

### Overview

The **penguinExplorer** R package provides both data and interactivity for exploring the *Palmer Penguins* dataset.  
It includes, a cleaned and well-documented `penguins` dataset, a Shiny app for interactive exploration and a full documentation and website built with **pkgdown**

This project demonstrates the integration of data communication, documentation and interactivity in R, turning static analysis into an engaging exploratory experience.

---

### Installation

To install and use the package directly from GitHub:

```{r}
install.packages("remotes", quiet = TRUE)
suppressPackageStartupMessages(
  remotes::install_github("ETC5523-2025/assignment-4-packages-and-shiny-apps-nishtabollavaram", quiet = TRUE)
)
library(penguinExplorer)

```

## Launch the Shiny app

Once installed, run the interactive app with:

```{r run_app, eval=FALSE}
run_app()
```

The interactive Shiny explorer allows you to: Select a specie, choose a numeric variables for X and Y axes, compare male and female penguins visually and download filtered data as a CSV.

## Dataset

The dataset contains information about body measurements for three species across the Palmer Archipelago, Antarctica.

```{r}
summary(penguins)



```

Variables include:
**species**, **island**, **sex**, **year**, **body_mass_g**	

## Acknowledgements
Data adapted from the palmerpenguins project.
Developed as part of the `ETC5523: Communicating with Data` unit at Monash University.

## License
This project is released under the **MIT License**.
