---
title: "getting-started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{getting-started}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(penguinExplorer)
```

### Introduction

Welcome! to **penguinExplorer**, an R package developed for **ETC5523:
Communicating with Data (Assignment 4)** by `Nishta Bollavaram`,
`Monash University`. 
This package demonstrates how to:
- Build an R package that includes data and documentation 
- Develop an interactive Shiny app to explore that data 
- Create a pkgdown website for public documentation

## Loading the Package
To install and load the package from GitHub:

```{r}
install.packages("remotes")
remotes::install_github("ETC5523-2025/assignment-4-packages-and-shiny-apps-nishtabollavaram")
library(penguinExplorer)


```

## The penguins Dataset

The package includes the **penguins** dataset (adapted from the Palmer Penguins dataset).
It contains information on body measurements of penguins from three species across different islands.

```{r, eval = FALSE}
head(penguins)
summary(penguins)


```

Variables include:
**species** — Penguin species (Adelie, Gentoo, Chinstrap)
**island** — Island where penguin was observed
**sex** — Male or female
**year** — Observation year
**body_mass_g** — Body mass in grams

## Launching the Shiny App

```{r}

run_app()


```

The interactive Shiny explorer allows you to:
- Select a species
- Choose numeric variables for X and Y axes
- Compare male and female penguins visually
- Download filtered data as a CSV

**Using the App**
The app has two tabs:
**Explore** — Displays an interactive scatterplot and summary table
**About** — Describes the dataset and app usage
The plot updates automatically when you change inputs.

**Summary Statistics**
species-level summaries can also be viewed using base R functions:

```{r, eval = FALSE}
penguins |>
dplyr::group_by(species) |>
dplyr::summarise(across(where(is.numeric), mean, na.rm = TRUE))


```

**Closing Notes**

This vignette provides an overview of the **penguinExplorer** package.
It demonstrates good R package structure, reproducible documentation, and communication through interactivity.
